---
title: "Bayesian Methods for CDH dataset"
author: 'Zanis Fang, UID: ZF2213'
date: "3/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
```


### Data cleaning

**Cases selection**
Death cases were removed because it would not reasonable to impute their follow-up data. After removal of death cases, "dead", "discharge", "SampleID", "dc_dead_age", "dc_alive_age" were also removed.

**Structural Imputation**
NA's in "gu_anomaly", "clp", "sk_anomalies", "bps", "ccam" can be imputed with 0 as long as "sys" is not NA. Since they added up to the number in "sys".

Making categorical variables factors for downstream anaylsis. "sys", "apgar_1", "apgar_5" remains continuous since they are ordinal.

```{r load_data}

# load data and impute several variables
cdh <- readxl::read_xlsx("./CDH_phenotypes.xlsx", sheet = 1, na = ".")

cdh[which(!is.na(cdh$sys)),] <- cdh %>% filter(!is.na(sys)) %>%
	# imputation, add up to "sys" column
	replace_na(list(gu_anomaly = 0, clp = 0,
									sk_anomalies = 0, bps = 0, ccam = 0))

# make categorical variables factors
for (i in 1:ncol(cdh)) {
	# turn numeric variables into factor variables
	if (length(unique(cdh[[i]])) <= 4) {
		cdh[[i]] <- as.factor(cdh[[i]])
	}
}

# make categorical variables ordered factors

# cdh <- cdh %>%
# 	mutate(apgar_1 = factor(apgar_1, ordered = TRUE),
# 				 apgar_5 = factor(apgar_5, ordered = TRUE),
# 				 sys = factor(sys, ordered = TRUE))

# cdh %>% skimr::skim()

# filter those alive
cdh_alive <- cdh %>% filter(dead == 0)

# data cleaning
cdh_iq_pca <- cdh_alive %>%
	# single value
	# select(-SampleID)
	# select(-SampleID, -dc_alive_age, -dc_dead_age)
	select(-dead, -discharge, -SampleID, -dc_dead_age, -dc_alive_age)
```

**Missing value imputation using EM algorithm**

Use expectation-maximization (EM) algorithm to impute missing values. Setting 3 underlying latent components.

```{r}
# estimate number of latent variables, run too long
# missMDA::estim_ncpFAMD(cdh_iq_pca, ncp.max = 4)
cdh_imputed_famd <- missMDA::imputeFAMD(cdh_iq_pca, ncp = 4)

# extract imputated dataframe
cdh_normalized <- cdh_imputed_famd$completeObs

```

**Scaling and centering the continuous variables**

Using square-root unit variance as scaling parameters.
Centering around zero.

```{r normalization}

# normalization

for (i in 1:ncol(cdh_normalized)) {
	if (!is.factor(cdh_normalized[[i]])) {
		cdh_normalized[[i]] <- as.numeric(pcaMethods::prep(cdh_normalized[[i]], scale = "pareto", center = TRUE))
	}
}

# cdh_normalized %>% skimr::skim()

# par(mfrow = c(1,2 ))
```

### Factor Analysis

FAMD (Factor Analysis for Mixed Data) is a principal component method dedicated to explore data with both continuous and categorical variables. It can be seen roughly as a mixed between PCA and MCA (multiple correspondence analysis).

Since the dataset contains a lot of categorical variables, it would be better to use factor analysis.

**Factor analysis results**

Eigenvalues:
```{r factor_analysis}

# factor analysis
cdh_famd <- FactoMineR::FAMD(cdh_normalized, ncp = 5, graph = FALSE)

cdh_famd$eig %>% knitr::kable()
# cdh_famd$ind$contrib %>% View
# cdh_famd$var$coord %>% View

cdh_famd_coord <- cdh_famd$ind$coord
```

Variables contribute most to the three dimensions are plotted and listed.

```{r top_variables}
cdh_famd$var$coord %>% as_tibble() %>% 
	arrange(desc(Dim.1)) %>% 
	mutate(num_row = row_number()) %>% 
	ggplot(aes(x = num_row, y = Dim.1)) +
	geom_point() +
	# reference line
	geom_vline(xintercept = 14.5) +
	labs(title = "Variable Loadings on 1st Dimension",
			 x = "Rank",
			 y = "Loadings on 1st Dimension") +
	theme_bw()

cdh_famd$var$coord %>% as_tibble() %>% 
	mutate(variable = rownames(cdh_famd$var$coord)) %>% 
	select(variable, Dim.1) %>% 
	top_n(14) %>% knitr::kable()

cdh_famd$var$coord %>% as_tibble() %>% 
	arrange(desc(Dim.2)) %>% 
	mutate(num_row = row_number()) %>% 
	ggplot(aes(x = num_row, y = Dim.2)) +
	geom_point() +
  # reference line
	geom_vline(xintercept = 13.5) +
	labs(title = "Variable Loadings on 2nd Dimension",
			 x = "Rank",
			 y = "Loadings on 2nd Dimension") +
	theme_bw()

cdh_famd$var$coord %>% as_tibble() %>% 
	mutate(variable = rownames(cdh_famd$var$coord)) %>% 
	select(variable, Dim.2) %>% 
	top_n(13) %>% knitr::kable()

cdh_famd$var$coord %>% as_tibble() %>% 
	arrange(desc(Dim.3)) %>% 
	mutate(num_row = row_number()) %>% 
	ggplot(aes(x = num_row, y = Dim.3)) +
	geom_point() +
  # reference line
	geom_vline(xintercept = 11.5) +
	labs(title = "Variable Loadings on 3rd Dimension",
			 x = "Rank",
			 y = "Loadings on 3rd Dimension") +
	theme_bw()

cdh_famd$var$coord %>% as_tibble() %>% 
	mutate(variable = rownames(cdh_famd$var$coord)) %>% 
	select(variable, Dim.3) %>% 
	top_n(11) %>% knitr::kable()


```


### Probabilistic graphical models

Gausian Mixture Models for K-means clustering. Three-centroid K-means using individual loadings on the three dimensions from FAMD analysis.

**Note that the K-means using factor loadings are quite different from K-means using original (imputed) dataset, the purpose of using the loadings is for interpretation**

```{r}
library(FPDclustering)
library(plotly)
fpdclust <- FPDC(cdh_famd_coord[, 1:4], k = 3, nf = 3, nu = 2)
#3,2,2(3)best nf cannot be larger than number of column, nf cannot be 1
Silh(fpdclust$probability)

kmeans_cdh <- tibble(
	cluster1 = fpdclust$probability[, 1],
	cluster2 = fpdclust$probability[, 2],
  cluster3 = fpdclust$probability[, 3],
	pc1 = cdh_famd_coord[,1],
	pc2 = cdh_famd_coord[,2],
	pc3 = cdh_famd_coord[,3],
	label = as.factor(fpdclust$label)
	) %>%
	# mutate(test = if_else(label == 1, cluster1, cluster2)) %>% 
	mutate(test = if_else(label == 1, cluster1,
												if_else(label == 2, cluster2, cluster3)))


# 2d plot
# kmeans_cdh %>%  ggplot(aes(x = pc1, y = pc2, color = label)) +
# 	geom_point(aes(alpha = test)) +
# 	xlim(c(-10, 10)) + 
# 	ylim(c(-10, 10))


# colors = c('#BF382A', '#0C4B8E')
plot_ly(kmeans_cdh,
				x = ~pc1, y = ~pc2, z = ~pc3, color = ~label, size = 1,
				marker = list(symbol = 'circle', sizemode = 'diameter'), sizes = 2) %>%
	add_markers() %>% 
  layout(scene = list(xaxis = list(title = 'NDD related',
  																 range = c(-20, 10)),
                     yaxis = list(title = 'Figures',
                     						 range = c(-10, 10)),
                     zaxis = list(title = 'CDH therapy',
                     						 range = c(-10, 5))))



```




```{r bayesian_pca, eval = FALSE, echo = FALSE}
library(pcaMethods)
set.seed(19)
# does not work for factor variables
resBPCA_cdh <- pcaMethods::pca(object = cdh_iq_pca,
															 scale = "pareto",
															 method = "bpca", center = TRUE, nPcs = 4)

b <- completeObs(resBPCA_cdh)


slplot(resBPCA_cdh, pcs = c(3,2))
Q2(resBPCA_cdh, fold = 5)

kEstimate(cdh_iq_pca, method = "bpca")
loadings(resBPCA_cdh)
scores(resBPCA_cdh)[,1:4]
as_tibble(scores(resBPCA_cdh)) %>%
	# mutate(group = as.factor(if_else(cdh_iq_pca$chd == 1,
	#																 if_else(cdh_iq_pca$sys > 1, 3, 2),
	#																 if_else(cdh_iq_pca$sys > 1, 1, 0)))) %>% 
	# filter(group %in% c(0)) %>% 
	ggplot(aes(x = PC2, y = PC1)) +
	geom_point(alpha = 0.5)
	# xlim(-1, 2) + ylim(-1.5, 1.5) +


variables <- rownames(loadings(resBPCA_cdh))
apca <- loadings(resBPCA_cdh) %>% as_tibble() %>% 
	mutate(name = variables)

apca %>% arrange(desc(abs(PC4)))
apca %>% arrange(desc(abs(PC3)))
apca %>% arrange(desc(abs(PC2)))
apca %>% arrange(desc(abs(PC1)))
```


```{r probabilistic_models, eval = FALSE, echo = FALSE}
library(FPDclustering)


pdclust <- PDclust(scores(resBPCA_cdh)[, 1:3], k = 2)
Silh(pdclust$probability)

fpdclust <- FPDC(scores(resBPCA_cdh)[, 1:3], k = 3, nf = 2, nu = 2)
#3,2,2(3)best nf cannot be larger than number of column, nf cannot be 1
Silh(fpdclust$probability)

tibble(
	cluster1 = fpdclust$probability[, 1],
	cluster2 = fpdclust$probability[, 2],
  cluster3 = fpdclust$probability[, 3],
	pc1 = scores(resBPCA_cdh)[,1],
	pc2 = scores(resBPCA_cdh)[,2],
	pc3 = scores(resBPCA_cdh)[,3],
	label = as.factor(fpdclust$label)
	) %>%
	# mutate(test = if_else(label == 1, cluster1, cluster2)) %>% 
	mutate(test = if_else(label == 1, cluster1,
												if_else(label == 2, cluster2, cluster3))) %>% 
	ggplot(aes(x = pc1, y = pc2, color = label)) +
	geom_point(aes(alpha = test))

tibble(
	prob = c(fpdclust$probability[, 1], 
					 fpdclust$probability[, 2],
					 fpdclust$probability[, 3]),
	pc1 = rep(scores(resBPCA_cdh)[,1], 3),
	pc2 = rep(scores(resBPCA_cdh)[,2], 3),
	pc3 = rep(scores(resBPCA_cdh)[,3], 3),
	pc = as.factor(c(rep(1, 471), rep(2, 471), rep(3, 471)))
) %>% ggplot(aes(x = pc1, y = pc2, color = pc)) +
	geom_point(aes(alpha = prob))


```


